---
title: "PA1_template"
author: "Carlos Ortega"
date: "November 8, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
# Loading Data
library(ggplot2)
library(plyr)
library(lattice)
```

## Importing Data
Reading data from activity.csv
```{r reading, cache=TRUE}
activity <- read.csv("activity.csv")
str(activity)
summary(activity)
```

#Processing the Data
Ordering the data by date and taking away the NA values
```{r processing, cache=TRUE}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime <- as.POSIXct(activity$date, format="%Y-%m-%d")
clean <- activity[!is.na(activity$steps),]
summary(clean)
```

## What is the average daily activity pattern?
#### 1. Calculate the total number of steps taken per day (ignore the missing values)
```{r calculate, cache=TRUE}
table <- aggregate(activity$steps ~ activity$date, FUN=sum)
colnames(table) <- c("Date", "Steps")
```
View total number of steps taken per day as a Histogram
```{r histogram, cache=TRUE}
hist(table$Steps, breaks = 5, xlab = "Steps", main = "Total Steps per Day")
```

```{r mean_median, cache=TRUE}
mean_steps <- as.integer(mean(table$Steps))
median_steps <- as.integer(median(table$Steps))
```
Mean and median of the total number of steps taken per day are `r mean_steps` steps and `r median_steps` steps, respectively.

## What is the average daily activity pattern?

#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r daily, cache=TRUE}
table <- ddply(clean, .(interval), summarize, Avg=mean(steps))

#we create a line plot with the number of steps per interval
p <- ggplot(table, aes(x=interval, y=Avg), xlab="Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")

#maximum steps by interval
max <- as.integer(max(table$Avg))

#maximum average number of steps
max_val <- table[table$Avg == max, 1]
```
The maximum number of steps for a 5-minute interval was `r max` steps.

The 5-minute interval which had the maximum number of steps was the `r max_val` interval.

## Imputing missing values

#### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r total_NAs, cache=TRUE}
#number of NAs in the original data set
total_nas <- nrow(activity) - nrow(clean)
```
The total number of NAs is `r total_nas`

```{r missing, cache=TRUE}
#Create the average number of steps per weekday and interval
table <- ddply(clean, .(interval, day), summarize, Avg=mean(steps))

#Dataset with all NAs
nadata <- activity[is.na(activity$steps),]

#merge na data with average weekday interval for substitution
newdata <- merge(nadata, table, by=c("interval", "day"))

#reorder the data in order to have the same format as the clean data
newdataTwo <- newdata[,c(6,4,1,2,5)]
colnames(newdataTwo) <- c("steps", "date", "interval", "day", "DateTime")

#merge the NA data (newdata) + non NA data (clean)
mergeVar <- rbind(clean, newdataTwo)

#Create sum of steps per date to compare with step 1
sumTable <- aggregate(mergeVar$steps ~ mergeVar$date, FUN=sum)
tableSum <- aggregate(activity$steps ~ activity$date, FUN=sum)
colnames(sumTable) <- c("Date", "Steps")

# mean
mean_missing <- as.integer(mean(sumTable$Steps))

#median
median_missing <- as.integer(median(sumTable$Steps))

#Create the histogram displaying the total steps per day

hist(sumTable$Steps, breaks = 5, xlab = "Steps", main = "Total Steps per Day with NAs Fixed", col = "Black")
hist(tableSum$Steps, breaks = 5, xlab = "Steps", main = "Total Steps per Day with NAs Fixed", col = "Grey", add = T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill = c("black", "grey"))
```

```{r calc_difference, cache=TRUE}
mean_diff <- mean_missing - mean_steps
median_diff <- median_missing - median_steps
```
The new mean of the imputed data is `r mean_missing` steps compared to the old mean of `r mean_steps` steps. That creates a difference of `r mean_diff` steps on average per day.

The new median of the imputed data is `r median_missing` steps compared to the old median of `r median_steps` steps. That creates a difference of `r median_diff` steps for the median.

## Activity Patterns Between Weekdays and Weekends

#### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

#### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r patterns, cache=TRUE}
mergeVar$DayCategory <- ifelse(mergeVar$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

#summarize data by interval and type o the day
interval <- ddply(mergeVar, .(interval, DayCategory), summarize, Avg=mean(steps))

#Plot
xyplot(Avg~interval|DayCategory, data=interval, type="l", layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day",
       ylab="Average Number of Steps", xlab="Interval")
```